---
title: "Quality control and analysis of IPF human references"
author: "Kolja becker and Sergio Picart-Armada"
date: "10 November, 2020"
output:
  html_document:
    toc: TRUE
    toc_float: true
    code_folding: hide
    df_print: paged
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(
  echo = TRUE, 
  message = FALSE, 
  error = FALSE, 
  warning = FALSE, 
  fig.height = 4)

```

# Introduction

Time when Rmd rendering started

```{r}

date()

```

Retrieve datasets:

* Kaminsky: `se.ipf.kaminsky`


```{r}

library(data.table)
library(SummarizedExperiment)
library(limma)
library(tidyverse)
library(magrittr)
library(ggplot2)
library(ggpmisc)
library(seqsetvis)
library(plotly)
library(pheatmap)


stopifnot(packageVersion("IST") >= "0.5.26")
library(IST)

# new config file
conf <- config::get()

# some helpers for GSEA/ORA
source("../00_helpers.R")

dir_out <- gsub("(^.*/)([^/]+)(/[^/]+$)", "\\2", conf$analysis_ipf_genesets)
if (!dir.exists(dir_out)) dir.create(dir_out)

```

# Load, PCA, Limma

```{r}
se.human <- readRDS(paste0("../", conf$interimdata_ipf_human))
list.sig <- readRDS(paste0("../", conf$interimdata_ipf_signatures_lfc))
list.sig.signif <- readRDS(paste0("../", conf$interimdata_ipf_signatures_lfc_signif))
meta.sig <- read.csv(paste0("../", conf$interimdata_ipf_signatures_meta))
list.orth <- IST::data.list.orth

df.kegg <- read.csv(paste0("../", conf$interimdata_genesets_kegg))
df.reactome <- read.csv(paste0("../", conf$interimdata_genesets_reactome))

# thresholds
th.logfc <- conf$threshold_dge_logfc
th.fdr <- conf$threshold_dge_fdr

# human signatures
sig.hsa <- subset(meta.sig, organism.name == "Human")$contrast.label
sig.hsa.ipf <- subset(meta.sig, organism.name == "Human" & contrast.type == "AnimalModel")$contrast.label
```

## Kaminsky dataset

```{r}

se.human

```

```{r}

sum.human <- summarise_dataset(
  se.human,
  return.interim = TRUE) 

# PCA plot kaminski
pca.human = sum.human$df.plot %>%
  arrange(is.medioid) %>%
  mutate(group = gsub('control', 'Control', group)) %>%
    ggplot() +
    # SPA: took out medioid as colour (all values are TRUE)
    # aes(x = PC1, y = PC2, fill = group, colour = is.medioid) +
    aes(x = PC1, y = PC2, fill = group) +
    geom_point(size=1, pch = 21) +
    scale_colour_manual(values = c("FALSE" = "gray70", "TRUE" = "black")) +
    scale_fill_manual(values = conf$col_ipf) +
    # labs(fill = "Group", colour = 'Medioid') +
    labs(fill = "Group") +
    gg_def

# save
ggsave(
  pca.human,
  file=paste0(dir_out, '/Fig_IPF_pca_kaminski.pdf'),
  width=80,
  height=50,
  units='mm')

```

```{r}

table(se.human$ist.class)

```

Age effect

```{r}

ggplot(sum.human$df.plot, aes(x = PC1, y = PC2, fill = age.ch1)) +
  geom_point(pch = 21) +
  scale_fill_gradient(low = "white", high = "indianred") +
  theme_bw()

# NOTE Kolja: Do we want this plot in the figures?

```



# Differential gene expression

```{r, get limma tables}

# get limma tables for human data
df.DE = lapply(list.sig[sig.hsa.ipf], 
               function(x) tibble::rownames_to_column(.data = x, 'ensemblID')) %>%
  bind_rows(.id = 'dataset') %>%
  mutate(sig = (adj.P.Val < th.fdr & abs(logFC) > th.logfc)) %>%
  data.table

```

## Violin plots

```{r, plots DE analysis}

# violin plot fold changes
violin.FC = df.DE %>%
  ggplot() +
    aes(x=dataset, y=logFC, fill=dataset) +
    geom_violin(lwd=0) +
    gg_def +
    theme(
      axis.text.x = element_text(angle=90, vjust=0.5, hjust=1),
      axis.title.x = element_blank(),
      legend.position = 'none')

# save
ggsave(
  violin.FC,
  file=paste0(dir_out, '/Fig_IPF_violin_FC.pdf'),
  height=50,
  width=30,
  units='mm'
)

# barplot DE genes
# Note Kolja: we could indicate how often each gene appears in the three
# 
# now there's only one signature with 4k DEGs approx
# barplot with a single bar
# 
# barplot.DE = df.DE[(sig)] %>%
#   add_count(ensemblID) %>%
#   ggplot() +
#   aes(x=dataset, fill=dataset) +
#     geom_bar(stat='count') +
#     gg_def +
#     theme(
#       axis.text.x = element_text(angle=90, vjust=0.5, hjust=1),
#       axis.title.x = element_blank(),
#       legend.position = 'none')
# 
# # save
# ggsave(
#   barplot.DE,
#   file=paste0(dir_out, '/Fig_IPF_barplot_DE.pdf'),
#   height=50,
#   width=30,
#   units='mm'
# )

```

## barplots

```{r, fig.height=1.5, fig.width=3}
dplyr::filter(meta.sig, contrast.type == "AnimalModel") %>%
    mutate(contrast.label = factor(contrast.label, levels = rev(conf$plot_sig_ipf_anmod))) %>%
    ggplot(aes(x = contrast.label, y = n.signif, fill = animal.model)) +
    geom_bar(stat = "identity", color = "gray30") +
    xlab("") +
    ylab("Number of significant genes") + 
    coord_flip() +
    scale_fill_manual(values = conf$col_ipf_mod, name = "Contrast type") +
    gg_def +
    theme(panel.grid.major.y = element_blank())

ggsave(paste0(dir_out, "/barplot_signatures_ipf_anmod_nsignif.pdf"))
```


## venn DEGs

```{r, venn diagram DEGs}

# venn diagram DE genes
# VENN diagram only has one circle now
# # 
# venn.DE = seqsetvis::ssvFeatureVenn(
#   df.DE[sig == TRUE] %>% split(x = .$ensemblID, f = .$dataset),
#   line_width = 0.5,
#   counts_txt_size = 1)+
#   theme(
#     legend.key.size=unit(0.5,"line"),
#     legend.position = 'right',
#     text=element_text(size=6)
#   )
# 
# # save
# ggsave(
#   venn.DE,
#   file=paste0(dir_out, '/Fig_IPF_venn_DE.pdf'),
#   height=40,
#   width=65,
#   units='mm'
# )

```

```{r, hypergeometric test}

# # DE genes to list
# groups = unique(df.DE$dataset)
# list.DE.all = lapply(groups, function(x)
#   df.DE[(sig)][dataset == x, ensemblID]) %>%
#   magrittr::set_names(paste0(groups, '_all'))
# list.DE.up = lapply(groups, function(x)
#   df.DE[(sig)][logFC > 0][dataset == x, ensemblID]) %>%
#   magrittr::set_names(paste0(groups, '_up'))
# list.DE.down = lapply(groups, function(x)
#   df.DE[(sig)][logFC < 0][dataset == x, ensemblID]) %>%
#   magrittr::set_names(paste0(groups, '_down'))
# 
# # all groups together
# list.DE = c(list.DE.all, list.DE.up, list.DE.down)
# 
# # hypergeometric test
# bkg = Reduce(intersect, 
#   list(rownames(data.kaminski), rownames(data.yang), rownames(data.meltzer)))
# res.hyper = compareSets(list.DE, bkgSet = bkg)

```

# Export main pathways

Combine KEGG and Reactome pathways

```{r}
# merge msig and kegg
df.pathway2gene <- bind_rows(
  data.frame(df.reactome, db = "Reactome"), 
  data.frame(df.kegg, db = "KEGG")
)

df.pathwaymeta <- add_count(df.pathway2gene, pathway) %>%
  dplyr::select(pathway, n, db) %>%
  unique 
```

Select known relevant pathways

```{r}
list.IPF.pathways.main <- c(
  "Extracellular matrix organization [Reactome]", 
  "Activation of matrix metalloproteinases [Reactome]", 
  "Collagen degradation [Reactome]", 
  "Collagen formation [Reactome]", 
  "VEGF signaling pathway [KEGG]", 
  "Toll like receptor cascades [Reactome]"
)

intersect(list.IPF.pathways.main, df.pathway2gene$pathway)
```

Other candidate pathways

```{r}
list.IPF.pathways.suppl <- c(
  "Organelle biogenesis and maintenance [Reactome]", 
  "Cilium assembly [Reactome]", 
  "Tak1 activates nfkb by phosphorylation and activation of ikks complex [Reactome]", 
  "Defensins [Reactome]", 
  "Anchoring of the basal body to the plasma membrane [Reactome]", 
  "Assembly of collagen fibrils and other multimeric structures [Reactome]", 
  "Ddx58 ifih1 mediated induction of interferon alpha beta [Reactome]", 
  "Collagen chain trimerization [Reactome]", 
  "Myd88 cascade initiated on plasma membrane [Reactome]", 
  "Trif ticam1 mediated tlr4 signaling [Reactome]", 
  "Toll like receptor 9 tlr9 cascade [Reactome]", 
  "Toll like receptor tlr1:tlr2 cascade [Reactome]", 
  "Interleukin 1 signaling [Reactome]", 
  "Toll like receptor 4 tlr4 cascade [Reactome]", 
  "Interleukin 1 family signaling [Reactome]", 
  "NOD-like receptor signaling pathway [KEGG]", 
  "Degradation of the extracellular matrix [Reactome]", 
  "Natural killer cell mediated cytotoxicity [KEGG]", 
  "ECM-receptor interaction [KEGG]", 
  "VEGF signaling pathway [KEGG]", 
  "Focal adhesion [KEGG]", 
  "PI3K-Akt signaling pathway [KEGG]", 
  "Regulation of apoptosis [Reactome]", 
  "Apoptosis [KEGG]"
)

intersect(list.IPF.pathways.suppl, df.pathway2gene$pathway)
```

Export pathways, tagging the ones in the main body

```{r}
# for the publication
df.export <- df.pathway2gene %>%
  mutate(in.story = pathway %in% list.IPF.pathways.main) %>% 
  dplyr::rename(gene.id = ensembl_gene_id, path.id = pathway) %>%
  add_count(path.id) %>%
  mutate(path.id = paste0(path.id, " (n=", n, ")"))
  

arrange(df.export, -n) %>% head

write.csv(df.export, file = paste0("../", conf$analysis_ipf_genesets), row.names = FALSE)
```


# Reproducibility

```{r}
date()
```

```{r}
sessionInfo()
```
