---
title: "Description of the human data and pathways in the NASH IST case study"
author: "Sergio Picart-Armada"
date: "7th December, 2020"
output:
  html_document:
    toc: TRUE
    toc_float: true
    code_folding: hide
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, 
                      error = FALSE, warning = FALSE, 
                      fig.height = 4)
```

# Introduction

Time when Rmd rendering started

```{r}
date()
```

Load data and libraries

```{r}
library(SummarizedExperiment)

library(data.table)
library(tidyverse)
library(magrittr)

library(ggplot2)
library(ggpmisc)
library(plotly)
library(pheatmap)

library(openxlsx)
library(clusterProfiler)

stopifnot(packageVersion("IST") >= "0.5.26")
library(IST)

# new config file
conf <- config::get()

source("../00_helpers.R")

```

```{r}
# get directory from config.yml file
dir_out <- gsub("(^.*/)([^/]+)(/[^/]+$)", "\\2", conf$analysis_nash_genesets)
if (!dir.exists(dir_out)) dir.create(dir_out)
```

# Load data

```{r}
se.human <- readRDS(paste0("../", conf$interimdata_nash_human))
list.sig <- readRDS(paste0("../", conf$interimdata_nash_signatures_lfc))
list.sig.signif <- readRDS(paste0("../", conf$interimdata_nash_signatures_lfc_signif))
meta.sig <- read.csv(paste0("../", conf$interimdata_nash_signatures_meta))
list.orth <- IST::data.list.orth

df.kegg <- read.csv(paste0("../", conf$interimdata_genesets_kegg))
df.reactome <- read.csv(paste0("../", conf$interimdata_genesets_reactome))

# thresholds
th.logfc <- conf$threshold_dge_logfc
th.fdr <- conf$threshold_dge_fdr

# human signatures
sig.hsa <- subset(meta.sig, organism.name == "Human")$contrast.label %>% as.character


```


# Human expression descriptive stats

```{r}

se.human

```

```{r}

sum.human <- summarise_dataset(
  se.human,
  return.interim = TRUE) 

```
Fibrosis stage

```{r}
ggplot(sum.human$df.plot, aes(x = PC1, y = PC2, fill = fibrosis_stage)) +
  geom_point(pch = 21) +
  # scale_colour_manual(values = c("FALSE" = "gray70", "TRUE" = "black")) +
  theme_bw()

ggsave(paste0(dir_out, "/pca_human.pdf"))
```

Numbers by phase

```{r}

table(se.human$fibrosis_stage)

```


Steatosis score

```{r}
ggplot(sum.human$df.plot, aes(x = PC1, y = PC2, fill = steatosis_score)) +
  geom_point(pch = 21) +
  # scale_colour_manual(values = c("FALSE" = "gray70", "TRUE" = "black")) +
  theme_bw()
```

Batch effect

```{r}

ggplot(sum.human$df.plot, aes(x = PC1, y = PC2, fill = Batch)) +
  geom_point(pch = 21) +
  # scale_fill_gradient(low = "white", high = "indianred") +
  theme_bw()

```

Age effect

```{r}

ggplot(sum.human$df.plot, aes(x = PC1, y = PC2, fill = age)) +
  geom_point(pch = 21) +
  scale_fill_gradient(low = "white", high = "indianred") +
  theme_bw()

```


# Human signatures descriptive stats

Distribution of logFCs

```{r, fig.width=7.5}
dplyr::bind_rows(list.sig.signif[sig.hsa], .id = "contrast") %>%
  dplyr::mutate(disease = ifelse(grepl("disease", contrast), "Disease", "Recovery")) %>%
  ggplot(aes(x = logFC)) +
  geom_histogram() +
  facet_wrap(disease~contrast, scales = "fixed", nrow = 2) +
  theme_bw()

ggsave(paste0(dir_out, "/logfc_histograms.pdf"))
```

```{r, fig.width=7.5}
dplyr::bind_rows(list.sig.signif[sig.hsa], .id = "contrast") %>%
  dplyr::mutate(mlog10FDR = -log10(adj.P.Val), 
                disease = ifelse(grepl("disease", contrast), "Disease", "Recovery")) %>%
  ggplot(aes(x = logFC, y = mlog10FDR)) +
  geom_point(size = .2) +
  facet_wrap(disease~contrast, scales = "fixed", nrow = 2) +
  theme_bw()

ggsave(paste0(dir_out, "/logfc_volcano.pdf"))
```


# General signature plots

## barplots

```{r, fig.height=1.5, fig.width=3}
dplyr::filter(meta.sig, contrast.label %in% conf$plot_sig_nash_anmod) %>%
    mutate(contrast.label = factor(contrast.label, levels = rev(conf$plot_sig_nash_anmod))) %>%
    ggplot(aes(x = contrast.label, y = n.signif, fill = animal.model)) +
    geom_bar(stat = "identity", color = "gray30") +
    xlab("") +
    ylab("Number of significant genes") + 
    coord_flip() +
    scale_fill_manual(values = conf$col_nash_mod, name = "Contrast type") +
    gg_def +
    theme(panel.grid.major.y = element_blank())

ggsave(paste0(dir_out, "/barplot_signatures_nash_anmod_nsignif.pdf"))
```

```{r, fig.height=1.5, fig.width=3}
dplyr::filter(meta.sig, contrast.label %in% conf$plot_sig_nash_trt) %>%
    mutate(contrast.label = factor(contrast.label, levels = rev(conf$plot_sig_nash_trt))) %>%
    ggplot(aes(x = contrast.label, y = n.signif, fill = animal.model)) +
    geom_bar(stat = "identity", color = "gray30") +
    xlab("") +
    ylab("Number of significant genes") + 
    coord_flip() +
    scale_fill_manual(values = conf$col_nash_mod, name = "Contrast type") +
    gg_def +
    theme(panel.grid.major.y = element_blank())

ggsave(paste0(dir_out, "/barplot_signatures_nash_trt_nsignif.pdf"))
```

# Gene set enrichment

Merge Reactome and KEGG

```{r}
# merge msig and kegg
df.pathway2gene <- bind_rows(
  data.frame(df.reactome, db = "Reactome"), 
  data.frame(df.kegg, db = "KEGG")
)

df.pathwaymeta <- add_count(df.pathway2gene, pathway) %>%
  dplyr::select(pathway, n, db) %>%
  unique 
```


## Knowledge-driven preselection


```{r}
# topics for selection
list.TA.regex <- c(
  "APOPTOSIS", 
  "IMMUN", 
  "FATTY.ACID", 
  "EXTRACELLULAR.MATRIX.ORGANIZATION", 
  "OXIDATIVE.STRESS",
  "PPAR"
)

list.path.regex <- lapply(setNames(list.TA.regex, list.TA.regex), grep, 
                          unique(df.pathway2gene$pathway), value = TRUE, ignore.case = TRUE)

list.path.regex

```
Metadata about the matches

```{r}
lapply(list.path.regex, function(x) subset(df.pathwaymeta, pathway %in% x))
```

For the publication, fix the long PPARalpha pathway. 
Due to code dependencies, this is the best place to do it. 
Now *Regulation of lipid metabolism by peroxisome proliferator activated receptor alpha PPARalpha [Reactome]* becomes *Regulation of lipid metabolism by PPARalpha [Reactome]*

```{r}
df.pathway2gene <- mutate(
    df.pathway2gene, 
    pathway = gsub("peroxisome proliferator activated receptor alpha ", "", pathway)
)

df.pathwaymeta <- mutate(
    df.pathwaymeta, 
    pathway = gsub("peroxisome proliferator activated receptor alpha ", "", pathway)
)
```


Pathway selection

```{r}
list.candidates <- c(
  "Apoptosis [KEGG]", 
  "Regulation of apoptosis [Reactome]", 
  "Cytokine signaling in immune system [Reactome]", 
  "Fatty acid metabolism [Reactome]", 
  "Fatty acid metabolism [KEGG]", 
  "Extracellular matrix organization [Reactome]", 
  "Oxidative stress induced senescence [Reactome]",
  "Regulation of lipid metabolism by PPARalpha [Reactome]", 
  "PPAR signaling pathway [KEGG]"
)

# all of them are exact matches in the gene sets
intersect(list.candidates, df.pathwaymeta$pathway)
```

Export table with those. 
For publication, shorten the PPARalpha pathway name (take out full name, keep acronym)

```{r}
# for the publication
df.export <- df.pathway2gene %>%
  mutate(in.story = pathway %in% list.candidates) %>% 
  dplyr::rename(gene.id = ensembl_gene_id, path.id = pathway) %>%
  add_count(path.id) %>%
  mutate(path.id = paste0(path.id, " (n=", n, ")"))
  

arrange(df.export, -n) %>% head

write.csv(df.export, file = paste0("../", conf$analysis_nash_genesets), row.names = FALSE)
```


# GSEA from clusterProfiler (with sign)

Double check lengths of fold change lists


```{r}
# lists with log FC 
list.gsea.up <- list.sig[sig.hsa] %>%
  lapply(function(x) {
    sort(setNames(x$logFC, rownames(x)), decreasing = TRUE)
  }) 

# export fold changes too
df.fc <- lapply(list.sig[sig.hsa], tibble::rownames_to_column, "gene.id") %>%
  dplyr::bind_rows()
dim(df.fc)
openxlsx::write.xlsx(df.fc, paste0(dir_out, "/df.export.logfc.xlsx"))


sapply(list.gsea.up, length)
```
Heads and tails of the list elements - sanity check

```{r}
lapply(list.gsea.up, function(x) c(head(x, 3), tail(x, 3))) 
```



Added more permutations and a seed

```{r}


# GSEA
res.gsea.fc.dir <- lapply(list.gsea.up, function(x)
  clusterProfiler::GSEA(
    x, 
    TERM2GENE=df.pathway2gene,
    minGSSize = 20,
    maxGSSize = 1200,
    pvalueCutoff = 1.1, 
    nPerm = 100000,
    pAdjustMethod='BH', 
    seed = TRUE) %>%
  .@result %>%
  data.table) %>%
  bind_rows(.id = 'dataset')


# ES and NES should always have the same sign
stopifnot(all(sign(res.gsea.fc.dir$enrichmentScore) == sign(res.gsea.fc.dir$NES)))

# save gsea results
openxlsx::write.xlsx(res.gsea.fc.dir, paste0(dir_out, "/df.export.gsea.logfc.dir.xlsx"))

# table
cat(list.candidates %in% res.gsea.fc.dir[p.adjust < 0.05, ID], sep='\n')

```

### Distribution of NES

```{r, fig.width=12}
mutate(res.gsea.fc.dir, signif = ifelse(qvalues < .05, "Significant", "Non-significant")) %>%
  ggplot(aes(x = NES)) +
  geom_vline(xintercept = 0) +
  geom_histogram() +
  facet_grid(signif~dataset) +
  theme_bw()

ggsave(paste0(dir_out, "/gsea_nes_histograms.pdf"))
```

### Number of hits

Total number of gene sets

```{r}
nrow(df.pathwaymeta)
```


```{r}
res.gsea.fc.dir %>%
  dplyr::filter(qvalues < .05) %>%
  add_count(dataset) %>%
  dplyr::select(dataset, n) %>% 
  unique
```


### Significant terms

At 5% FDR:

```{r, fig.width=12, fig.height=5}
# heatmap 
res.gsea.fc.dir[ID %in% list.candidates] %>%
         mutate(log.fdr = ifelse(qvalues < 0.05, -log10(qvalues), NA)) %>%
  ggplot(aes(x=dataset, y=ID, color=log.fdr)) +
  geom_point() +
  scale_color_gradient(low = "orange", high = "red") +
  theme(
    axis.text.x=element_text(angle=90, hjust=1, vjust=0.5),
    axis.title=element_blank())

ggsave(paste0(dir_out, "/gsea_significant_hits_FDR005.pdf"))
```

At 10% FDR:

```{r, fig.width=12, fig.height=5}
# heatmap 
ggplot(res.gsea.fc.dir[ID %in% list.candidates] %>%
         mutate(log.fdr=ifelse(qvalues < 0.1, -log10(qvalues), NA))) +
  aes(x=dataset, y=ID, color=log.fdr) +
  geom_point() +
  scale_color_gradient(low = "orange", high = "red") +
  theme(
    axis.text.x=element_text(angle=90, hjust=1, vjust=0.5),
    axis.title=element_blank())
```

Heatmap NES

```{r, fig.width=12, fig.height=5}
# heatmap 
ggplot(res.gsea.fc.dir[ID %in% list.candidates] %>%
         mutate(signif = qvalues < 0.1)) +
  aes(x=dataset, y=ID, fill = NES, colour = signif) +
  geom_point(pch = 21) +
  scale_colour_manual(values = c("TRUE" = "black", "FALSE" = "white")) +
  scale_fill_gradient2(low = conf$col_gsea["low"], mid = conf$col_gsea["mid"], high = conf$col_gsea["high"]) +
  theme(
    axis.text.x=element_text(angle=90, hjust=1, vjust=0.5),
    axis.title=element_blank())

ggsave(paste0(dir_out, "/gsea_nes.pdf"))
```




## Zip files

Files not zipped for now

```{r}
# zip.file <- paste0(dir_out, "/export_plots.zip")
# list.files <- paste0(dir_out, "/", list.files(dir_out, recursive = TRUE))
# 
# unlink(zip.file)
# zip(zipfile = zip.file, files = list.files)
```


# Reproducibility

```{r}
date()
```

```{r}
sessionInfo()
```
