---
title: "Adding null model to IST in NASH"
author: "Sergi Picart and Kolja Becker"
date: "10/02/2021"
output:
  html_document:
    toc: TRUE
    toc_float: true
    code_folding: hide
    df_print: paged
---

# Introduction

Time when Rmd rendering started

```{r}
date()
```

Load data and libraries

```{r, message=FALSE, warning=FALSE}
library(SummarizedExperiment)

library(tidyverse)
library(magrittr)

library(ggplot2)
library(ggpmisc)
library(plotly)
library(pheatmap)

stopifnot(packageVersion("IST") >= "0.5.26")
library(IST)

library(parallel)

source("../00_helpers.R")

# new config file
conf <- config::get()

set.seed(42)
```

```{r}
# get directory from config.yml file
# dir_out <- gsub("(^.*/)([^/]+)(/[^/]+$)", "\\2", conf$analysis_nash_genesets)
dir_out <- "3_IST_disease_output/"
if (!dir.exists(dir_out)) dir.create(dir_out)
```


# Load relevant data

We need

* Signature list
* Signature metadata
* Orthology mapping
* Pathways

```{r}
# remove small (<100DEG) signatures
list.sig <- readRDS(paste0("../", conf$interimdata_nash_signatures_lfc))
list.sig.signif <- readRDS(paste0("../", conf$interimdata_nash_signatures_lfc_signif))
meta.sig <- read.csv(paste0("../", conf$interimdata_nash_signatures_meta), row.names = 1, stringsAsFactors = FALSE)

# signature metadata for shiny app
meta.sig$sig.class <- ifelse(
    grepl("F4", meta.sig$sig.id), "positivecontrol", 
    ifelse(meta.sig$organism.name == "Mouse", "default", "other")
)

list.orth <- IST::data.list.orth
df.path <- read.csv(paste0("../", conf$analysis_nash_genesets), stringsAsFactors = FALSE)
df.path.bin <- filter(df.path, in.story) %>% select(path.id, gene.id)
se.human <- readRDS(paste0("../", conf$interimdata_nash_human))

# labels for gene heatmaps
vec.ensembl2symbolonly <- readRDS(paste0("../", conf$interimdata_genelabels))

# pathway metadata for shiny app
df.path.meta <- data.frame(
    path.id = unique(df.path.bin$path.id), 
    path.class = "default", 
    stringsAsFactors = FALSE
)

sig.anmod <- conf$analysis_nash_signaturelist_anmod
sig.trt <- conf$analysis_nash_signaturelist_trt
```

Separate animal model and treatment signatures


```{r}
meta.sig.anmod <- meta.sig[sig.anmod, ]
meta.sig.trt <- meta.sig[sig.trt, ]
```


# Null model

### Permuted datasets

```{r}
# set.seed(1)
# permute_class <- function(se, seed, ...) {
#   set.seed(seed)
#   se$ist.class %<>% sample
#   se
# }
# 
# n.perm.se <- 10
# list.se.perm <- lapply(
#   seq_len(n.perm.se), 
#   permute_class, 
#   se = se.human
# ) %>% 
#   set_names(paste0("PERM", formatC(seq_len(n.perm.se), width = 2, flag = "0")))
```

Double check the class was permuted

```{r}
# sapply(list.se.perm, extract2, "ist.class") %>% as.data.frame
```

### Surrogated datasets

```{r}
# set.seed(1)
# surrogate_class <- function(se, prop = .8, seed, ...) {
#   set.seed(seed)
#   
#   ids <- caret::createDataPartition(se$ist.class, p = prop)[[1]]
#   se[, ids]
# }
# 
# list.se.surrogate <- lapply(
#   seq_len(n.perm.se), 
#   surrogate_class, 
#   se = se.main, 
#   prop = .8
# ) %>% 
#   set_names(paste0("SUR", formatC(seq_len(n.perm.se), width = 2, flag = "0")))
```


## Gene sets

### Permuted gene sets

Draft genes from the non-DE bucket

```{r}
# permute_pathway <- function(df, universe, ...) {
#   genes <- sample(universe, size = nrow(df), replace = FALSE)
#   
#   data.frame(path.id = df$path.id[1], gene.id = genes)
# }
# 
# set.seed(1)
# df.path.perm <- plyr::ddply(
#   df.path.bin, "path.id", permute_pathway, universe = de.not
# )
```

### Surrogated gene sets

Draft genes from the non-DE bucket

```{r}
# surrogate_pathway <- function(df, prop, ...) {
#   n.genes <- nrow(df)
#   
#   inds.keep <- sample(n.genes, size = prop*n.genes)
#   df[inds.keep, , drop = FALSE]
# }
# 
# set.seed(1)
# df.path.surrogate <- plyr::ddply(
#   df.path.bin, "path.id", surrogate_pathway, prop = .8
# )
```

## Signatures

### Permuted signatures

```{r}
# # permute a single signature
# permute_signature <- function(df.orig, df.signif, ...) {
#   n.resample <- nrow(df.signif)
#   genes.resample <- sample(rownames(df.orig), n.resample)
#   set_rownames(df.signif, genes.resample)
# }

# permutes a collection of signatures in a coordinated way
# (i.e. keeping a consistent mapping all the time)
# ASSUMES THE SIGNATURES HAVE A COMMON BACKGROUND
permute_signatures <- function(list.orig, list.signif, ...) {
  # background
  gene.universe <- lapply(list.orig, rownames) %>% unlist %>% unique
  # gene shuffling
  v.map <- setNames(gene.universe, sample(gene.universe))
  
  lapply(list.signif, function(df) {
    rn <- rownames(df)
    rn.new <- v.map[rn]
    rownames(df) <- rn.new
    
    df
  })
}

list.sig.bystudy <- split(list.sig, meta.sig$study.id)
list.sig.signif.bystudy <- split(list.sig.signif, meta.sig$study.id)

set.seed(1)
# for testing
# n.perm.sig <- 10
# for paper
n.perm.sig <- 100

list.list.sig.perm <- lapply(
  seq_len(n.perm.sig),
  function(dummy) {
    
    list.perm <- mapply(
      permute_signatures, 
      list.orig = list.sig.bystudy, 
      list.signif = list.sig.signif.bystudy, 
      SIMPLIFY = FALSE) 
    
    reduce(list.perm, c)[names(list.sig)]
    # unlist(use.names = TRUE, recursive = FALSE)
    # mapply(
    #   permute_signature, 
    #   df.orig = list.sig, 
    #   df.signif = list.sig.signif, 
    #   SIMPLIFY = FALSE
    # )
  }
) %>%
  set_names(paste0("PERM", formatC(seq_len(n.perm.sig), 
                                   width = floor(log10(n.perm.sig)) + 1, 
                                   flag = "0")))


list.list.sig.perm.anmod <- lapply(list.list.sig.perm, extract, sig.anmod)
list.list.sig.perm.trt <- lapply(list.list.sig.perm, extract, sig.trt)
```


```{r}
# pryr::object_size(list.list.sig.perm)
```


### Surrogated signatures

Must start from original signatures, before subsetting for DE genes!

```{r}
# # surrogate_pathway will also work here
# set.seed(1)
# 
# list.list.sig.surrogate <- lapply(
#   seq_len(n.perm.sig),
#   function(dummy) {
#     mapply(
#       surrogate_pathway, 
#       df = list.sig.signif, 
#       MoreArgs = list(prop = .8), 
#       SIMPLIFY = FALSE
#     )
#   }
# ) %>%
#   set_names(paste0("SUR", formatC(seq_len(n.perm.sig), width = 3, flag = "0")))
  
```

Size of a signatures object

```{r}
# pryr::object_size(list.list.sig.perm)
```


# Fit signatures object

Number of signatures

```{r}
length(list.sig)
```

Size of metadata table

```{r}
dim(meta.sig)
```

Signatures:

```{r}
sapply(list.sig.signif, nrow) %>% data.frame(n.significant = .)
```

## Animal models

```{r}
ist.sig.anmod <- define(
  "ist.signatures", 
  list.sig = list.sig.signif[sig.anmod],
  tab.meta = meta.sig[sig.anmod, ], 
  org.to = "hsapiens", 
  list.mapping = list.orth
) %>% fit
```

## Treatments

```{r}
ist.sig.trt <- define(
  "ist.signatures", 
  list.sig = list.sig.signif[sig.trt],
  tab.meta = meta.sig[sig.trt, ], 
  org.to = "hsapiens", 
  list.mapping = list.orth
) %>% fit
```

# Fit pathways objects

```{r}
list.path <- df.path.bin$path.id %>% unique
```

## Loop over human references

```{r, results='hide', message=FALSE, warning=FALSE}
# file.istpath <- paste0(dir.main, "/01_permphenos_list.istpath.rds")

# # permuted phenos
# if (!file.exists(file.istpath)) {
#   # this takes a while (~20min?)
#   list.istpath <- lapply(
#     c(list(Main = se.main), list.se.perm), 
#     function(se) {
#       define(
#           "ist.pathways",
#           X = t(assay(se)), 
#           y = se$ist.class, 
#           org.to = "hsapiens", 
#           id.bin = rep(TRUE, ncol(se)), 
#           id.oc = se$ist.class %in% c(1), 
#           pathways.table.bin = df.path.bin
#       ) %>% fit 
#     }
#   )
#   
#   # object weighs 1.7GB - can we save it?
#   # 800MB in disk - okay
#   saveRDS(list.istpath, file.istpath)  
# } else {
#   list.istpath <- readRDS(file.istpath)
# }

# original data
ist.path <- define(
    "ist.pathways",
    X = t(assay(se.human)), 
    y = se.human$ist.class, 
    org.to = "hsapiens", 
    id.bin = se.human$fibrosis_stage %in% c(0, 4),
    id.oc = se.human$ist.class %in% c(1), 
    pathways.table.bin = df.path.bin, 
    pathways.meta = df.path.meta
) %>% fit 

# # permuted GS
# ist.path <- define(
#     "ist.pathways",
#     X = t(assay(se.human)), 
#     y = se.human$ist.class, 
#     org.to = "hsapiens", 
#     id.bin = rep(TRUE, ncol(se.human)), 
#     id.oc = se.human$ist.class %in% c(1), 
#     pathways.table.bin = df.path.bin
# ) 
```

# Signature and pathway selections

Graphical parameters for heatmaps

```{r}
args.ph <- default.args.pheatmap("pathwaymap")
args.ph$cluster_cols <- TRUE
args.ph$cluster_rows <- TRUE

args.ph2 <- args.ph
args.ph2$cluster_rows <- FALSE

args.ph3 <- args.ph
args.ph3$cluster_cols <- FALSE

args.ph4 <- args.ph
args.ph4$cluster_cols <- FALSE
args.ph4$cluster_rows <- FALSE

# dont plot heatmap (silent)
args.ph5 <- args.ph4
args.ph5$silent <- TRUE

args.phnotext <- args.ph
args.phnotext$fontsize_number <- 0

args.phnotext2 <- args.phnotext
args.phnotext2$cluster_rows <- FALSE

vars.heatmap <- c("n.signif", "organism.name")
```


# Animal models

## Regular pathway heatmap

```{r}
ist.res.anmod <- define(
  "ist.results", 
  ist.signatures = ist.sig.anmod, 
  ist.pathways = ist.path, 
  id.ref = se.human$ist.class == -1, 
  id.ist = se.human$ist.class == 1,
  group = se.human$class, 
  vec.gene2label = vec.ensembl2symbolonly
) %>% fit
```

```{r, fig.height=6, fig.width=9}
mat.pathwaymap.anmod <- plot.ist.pathwaymaps(
  ist.res.anmod, 
  type = "pheatmap", args.pheatmap = args.ph4, 
  vars.meta.sig = c("animal.model", "organism.name", "n.signif"))
```

## Regular genemaps

Whole genemaps (limit to 500 genes... otherwise adobe text search breaks)

```{r, fig.width=22, fig.height=4}
save.genemaps.pdf(
    ist.results = ist.res.anmod, 
    dir.out = paste0(dir_out, "/genemaps-anmod-full"),
    # id.paths = v.gsmain, 
    # sig.ids = c(sig.main, sig.hsaF4), 
    args.pheatmap = args.phnotext, 
    vars.meta.sig = vars.heatmap, 
    cex.width = 10, 
    cex.height = 1.2,
    main.width = 120, 
    max.genes = 500
)
```

Trimmed genemaps

```{r, fig.width=16, fig.height=4}
save.genemaps.pdf(
    ist.results = ist.res.anmod, 
    dir.out = paste0(dir_out, "/genemaps-anmod-top50-tall"),
    # id.paths = v.gsmain, 
    # sig.ids = c(sig.main, sig.hsaF4), 
    args.pheatmap = args.phnotext, 
    vars.meta.sig = vars.heatmap, 
    cex.width = 1.5, 
    cex.height = 1.2,
    main.width = 120, 
    max.genes = 50
)
```

Trimmed genemaps

```{r, fig.width=16, fig.height=3}
save.genemaps.pdf(
    ist.results = ist.res.anmod, 
    dir.out = paste0(dir_out, "/genemaps-anmod-top50-short"),
    # id.paths = v.gsmain, 
    # sig.ids = c(sig.main, sig.hsaF4), 
    args.pheatmap = args.phnotext, 
    vars.meta.sig = vars.heatmap, 
    cex.width = 1.5, 
    cex.height = .5,
    main.width = 120, 
    max.genes = 50
)
```

## Recap when random signatures?


```{r}
get_ist_results <- function(se, ist.path, ist.sig, class.ref = "Stage 4", class.ist = "Stage 0") {
  define(
      "ist.results", 
      ist.signatures = ist.sig, 
      ist.pathways = ist.path, 
      group = se$class,
      id.ref = se$class == class.ref, 
      id.ist = se$class == class.ist,
      vec.gene2label = IST::vec.ensembl2symbol
    ) %>% fit
}
```


```{r}
get_ist_arrays <- function(se, ist.path, list.sig, meta.sig, 
                           list.mapping = list.orth, vec.gene2label = vec.ensembl2symbol,
                           class.ref = "Stage 4", class.ist = "Stage 0", 
                           vars.heatmap = c("n.signif", "organism.name"), args.pheatmap = args.ph5) {
  # get ist signatures object
  ist.sig <- IST::fit(IST::define(
      "ist.signatures", 
      list.sig = list.sig,
      tab.meta = meta.sig, 
      org.to = "hsapiens", 
      list.mapping = list.mapping
    )
  )
  # try({
    
    # get ist results object
  ist.res <- IST::fit(IST::define(
      "ist.results",
      ist.signatures = ist.sig,
      ist.pathways = ist.path,
      group = se$class,
      id.ref = se$class == class.ref,
      id.ist = se$class == class.ist,
      vec.gene2label = vec.gene2label
    )
  )
  # })
  
  # 
  # export main heatmap
  ist.plt <- IST::plot.ist.pathwaymaps(
    ist.res,
    type = "pheatmap",
    args.pheatmap = args.pheatmap,
    vars.meta.sig = vars.heatmap
  )

  ist.plt$plot.data$data.wide
}
```


```{r, results='hide', message=FALSE, fig.keep="last"}
list.perm.anmod <- parallel::mclapply(
  X = list.list.sig.perm.anmod, 
  FUN = get_ist_arrays,
  # MoreArgs = list(
    meta.sig = meta.sig.anmod, 
    ist.path = ist.path, 
    se = se.human,
    class.ref = "Stage 4", 
    class.ist = "Stage 0", 
    args.pheatmap = args.ph5, 
    vec.gene2label = IST::vec.ensembl2symbol, 
    list.mapping = list.orth,
  # ), 
  mc.cores = 1
  # mc.preschedule = TRUE, 
  # mc.cleanup = TRUE
) 

df.orig.anmod <- mat.pathwaymap.anmod$plot.data$data.long
mat.orig.anmod <- mat.pathwaymap.anmod$plot.data$data.wide
list.perm.sigs.anmod <- mat.orig.anmod %>% rownames

# fills with 0s missing rows
# takes order from full.rownames
fill_mat_missing_rows <- function(mat, full.rownames) {
  rownames.missing <- setdiff(full.rownames, rownames(mat))
  n.missing <- length(rownames.missing)
  
  # new matrix
  matdiff <- matrix(nrow = n.missing, ncol = ncol(mat), data = 0)
  rownames(matdiff) <- rownames.missing
  colnames(matdiff) <- colnames(mat)
  
  # to ensure order is consistent
  rbind(mat, matdiff)[full.rownames, , drop = FALSE]
}

array.perm.anmod <- plyr::laply(list.perm.anmod, fill_mat_missing_rows, full.rownames = list.perm.sigs.anmod)
dimnames(array.perm.anmod)[[1]] <- names(list.perm.anmod)
```

```{r}
# empirical p-value
# p = (r+1)/(n+1)
mat.p.anmod <- sweep(array.perm.anmod, c(2, 3), mat.orig.anmod, ">") %>%
  apply(c(2, 3), function(row) (sum(row) + 1)/(length(row) + 1))
mat.orig.signif.anmod <- mat.orig.anmod
mat.orig.signif.anmod[mat.p.anmod > .05] <- NA
  
df.perm.distrib.anmod <- plyr::adply(array.perm.anmod, c(2, 3), data.frame) %>%
  rename(sig.id = X1, pathway = X2, null.delta = piece)

```

P-values

```{r}
signif(mat.p.anmod, 2) %>% as.data.frame
```
Melt p-values table

```{r}
df.perm.p.anmod <- as.data.frame(mat.p.anmod) %>%
  rownames_to_column("sig.id") %>%
  tidyr::pivot_longer(-sig.id, values_to = "p.value", names_to = "pathway")
```

```{r}
df.perm.plot.anmod <- left_join(df.orig.anmod, df.perm.p.anmod) %>%
  mutate(signif = p.value < .05, signif.range = as.ordered(cut(p.value, c(0, .05, .2, 1))))
  # mutate(signif = p.value < .05)
  # mutate(significance = ifelse(p.value < .05, "p<0.05", "p>=.05"))
```


P-values, added on top of original recap

```{r, fig.height=6, fig.width=9}
plt.p <- do.call(
  pheatmap, 
  c(list(mat = mat.orig.signif.anmod, 
         color = heatmaps.pal(), 
         breaks = sym.breaks(mat.orig.anmod)
         ), args.ph4)
)

ggsave(paste0(dir_out, "/IST_NASH_anmod_withpvalue.png"), plot = plt.p)
```

Plot recap with significance (GSEA-like style)

```{r, fig.width=3, fig.height=2.25}

mutate(df.perm.plot.anmod, sig.id = factor(sig.id, levels = sig.anmod)) %>%
  ggplot(aes(y=sig.id, x=pathway, fill = total.delta.percent, 
           label = format(total.delta.percent, digits=0), color=signif.range)) +
  geom_tile(width=0.9, height=0.9, size=0.15) +
  ggplot2::geom_text(size = 1.1, color='black') +
  scale_colour_manual(values = c("black", "gray80", "white"), name = "Significance") +
  scale_fill_gradient2(low = "deepskyblue3", mid = "white", high = "darkorange1", name = "Recapitulation [%]", breaks = scales::pretty_breaks(n = 7)) +
  gg_def +
  # fix aspect ratio
  coord_fixed() +
  # fix legend
  guides(color = guide_legend(order = 2, override.aes = list(fill = "white")), 
         fill = guide_colorbar(order = 1)) +
  theme(
      # play with adding margins so that legend fits in plot
      legend.margin = margin(t = 0, r = 0, b = -.25, l = 0, unit = "cm"),
      plot.margin = margin(t = 1, r = .2, b = .2, l = .2, unit = "cm"),
      panel.grid = element_blank(),
    axis.line=element_blank(),
    panel.border=element_blank(),
    axis.ticks=element_blank(),
    text = element_text(size = 5, family = "Helvetica"),
    legend.key.size = unit(0.7,"line"),
    axis.text.x=element_text(angle=90, hjust=1, vjust=0.5),
    axis.title=element_blank())

ggsave(file = paste0(dir_out, '/main_nash_anmod_heatmap_extended.pdf'), width=80, height=70, units='mm')
```

Plot pathway selection

```{r, fig.width=3, fig.height=2}
subset(df.perm.plot.anmod, pathway %in% conf$analysis_nash_pathwayselection) %>%
  mutate(sig.id = factor(sig.id, levels = sig.anmod), 
         pathway = factor(pathway, levels = conf$analysis_nash_pathwayselection)) %>%
  ggplot(aes(y=sig.id, x=pathway, fill = total.delta.percent, 
           label = format(total.delta.percent, digits=0), color=signif.range)) +
  geom_tile(width=0.9, height=0.9, size=0.15) +
  ggplot2::geom_text(size = 1.1, color='black') +
  scale_colour_manual(values = c("black", "gray80", "white"), name = "Significance") +
  scale_fill_gradient2(low = "deepskyblue3", mid = "white", high = "darkorange1", name = "Recapitulation [%]", breaks = scales::pretty_breaks(n = 5)) +
  gg_def +
    # fix aspect ratio
  coord_fixed() +
  # fix legend
  guides(color = guide_legend(order = 2, override.aes = list(fill = "white")), 
         fill = guide_colorbar(order = 1)) +
  theme(
      # play with adding margins so that legend fits in plot
      legend.margin = margin(t = 0, r = 0, b = -.25, l = 0, unit = "cm"),
      plot.margin = margin(t = 1, r = .2, b = .2, l = .2, unit = "cm"),
      panel.grid = element_blank(),
    axis.line=element_blank(),
    panel.border=element_blank(),
    axis.ticks=element_blank(),
    text = element_text(size = 5, family = "Helvetica"),
    legend.key.size = unit(0.7,"line"),
    axis.text.x=element_text(angle=90, hjust=1, vjust=0.5),
    axis.title=element_blank())

ggsave(file = paste0(dir_out, '/main_nash_anmod_heatmap.pdf'), width=75, height=75, units='mm')
```

Plot distribution histograms

```{r, fig.width=12, fig.height=6}
list.path.plot <- list.path

ggplot() +
  geom_vline(xintercept = 0, colour = "gray50") +
  geom_vline(aes(xintercept = total.delta.percent, colour = signif.range), 
             lwd = .5, lty = 5, 
             data = subset(df.perm.plot.anmod, pathway %in% list.path.plot)) +
  geom_histogram(aes(x = null.delta), 
                 data = subset(df.perm.distrib.anmod, pathway %in% list.path.plot)) +
  facet_grid(sig.id~pathway, scales = "free") +
  scale_color_manual(values = c("indianred1", "darkgoldenrod2", "gray70")) +
  xlab("Recapitulation (null distribution, and observed statistic in dashed line)") +
  ylab("Counts (null distribution)") +
  theme_bw() +
  theme(strip.text.y = element_text(angle = 0), 
        strip.text.x = element_text(angle = 90))

ggsave(file = paste0(dir_out, '/main_nash_anmod_histograms.pdf'))
```


## Export data objects

```{r}
saveRDS(ist.res.anmod, paste0(dir_out, "/ist_res_anmod.rds"))
save(mat.orig.anmod, mat.p.anmod, array.perm.anmod, df.orig.anmod, df.perm.distrib.anmod, 
     df.perm.plot.anmod, file = paste0(dir_out, "/null_models_anmod.RData"))
```


# Treatments

## Regular heatmap

```{r}
ist.res.trt <- define(
  "ist.results", 
  ist.signatures = ist.sig.trt, 
  ist.pathways = ist.path, 
  id.ref = se.human$ist.class == 1, 
  id.ist = se.human$ist.class == -1,
  group = se.human$class, 
  vec.gene2label = vec.ensembl2symbolonly
) %>% fit
```

```{r, fig.height=6, fig.width=9}
mat.pathwaymap.trt <- plot.ist.pathwaymaps(
  ist.res.trt, 
  type = "pheatmap", args.pheatmap = args.ph4, 
  vars.meta.sig = c("animal.model", "organism.name", "n.signif"))
```

## Regular genemaps

Whole genemaps (limit to 500 genes... otherwise adobe text search breaks)

```{r, fig.width=22, fig.height=4}
save.genemaps.pdf(
    ist.results = ist.res.trt, 
    dir.out = paste0(dir_out, "/genemaps-trt-full"),
    # id.paths = v.gsmain, 
    # sig.ids = c(sig.main, sig.hsaF4), 
    args.pheatmap = args.phnotext, 
    vars.meta.sig = vars.heatmap, 
    cex.width = 2.5, 
    cex.height = 1.2,
    main.width = 120, 
    max.genes = 500
)
```

Trimmed genemaps

```{r, fig.width=16, fig.height=4}
save.genemaps.pdf(
    ist.results = ist.res.trt, 
    dir.out = paste0(dir_out, "/genemaps-trt-top50-tall"),
    # id.paths = v.gsmain, 
    # sig.ids = c(sig.main, sig.hsaF4), 
    args.pheatmap = args.phnotext, 
    vars.meta.sig = vars.heatmap, 
    cex.width = 1.5, 
    cex.height = 1.2,
    main.width = 120, 
    max.genes = 50
)
```

```{r, fig.width=16, fig.height=4}
save.genemaps.pdf(
    ist.results = ist.res.trt, 
    dir.out = paste0(dir_out, "/genemaps-trt-top50-short"),
    # id.paths = v.gsmain, 
    # sig.ids = c(sig.main, sig.hsaF4), 
    args.pheatmap = args.phnotext, 
    vars.meta.sig = vars.heatmap, 
    cex.width = 1.5, 
    cex.height = .5,
    main.width = 120, 
    max.genes = 50
)
```

## Recap when random signatures?


```{r, results='hide', message=FALSE, fig.keep="last"}
list.perm.trt <- parallel::mclapply(
  X = list.list.sig.perm.trt, 
  FUN = get_ist_arrays,
  # MoreArgs = list(
    meta.sig = meta.sig.trt, 
    ist.path = ist.path, 
    se = se.human,
    class.ref = "Stage 0", 
    class.ist = "Stage 4", 
    args.pheatmap = args.ph5, 
    vec.gene2label = IST::vec.ensembl2symbol, 
    list.mapping = list.orth,
  # ), 
  mc.cores = 1
  # mc.preschedule = TRUE, 
  # mc.cleanup = TRUE
) 

df.orig.trt <- mat.pathwaymap.trt$plot.data$data.long
mat.orig.trt <- mat.pathwaymap.trt$plot.data$data.wide
list.perm.sigs.trt <- mat.orig.trt %>% rownames

# fills with 0s missing rows
# takes order from full.rownames
fill_mat_missing_rows <- function(mat, full.rownames) {
  rownames.missing <- setdiff(full.rownames, rownames(mat))
  n.missing <- length(rownames.missing)
  
  # new matrix
  matdiff <- matrix(nrow = n.missing, ncol = ncol(mat), data = 0)
  rownames(matdiff) <- rownames.missing
  colnames(matdiff) <- colnames(mat)
  
  # to ensure order is consistent
  rbind(mat, matdiff)[full.rownames, , drop = FALSE]
}

array.perm.trt <- plyr::laply(list.perm.trt, fill_mat_missing_rows, full.rownames = list.perm.sigs.trt)
dimnames(array.perm.trt)[[1]] <- names(list.perm.trt)
```

```{r}
# empirical p-value
# p = (r+1)/(n+1)
mat.p.trt <- sweep(array.perm.trt, c(2, 3), mat.orig.trt, ">") %>%
  apply(c(2, 3), function(row) (sum(row) + 1)/(length(row) + 1))
mat.orig.signif.trt <- mat.orig.trt
mat.orig.signif.trt[mat.p.trt > .05] <- NA
  
df.perm.distrib.trt <- plyr::adply(array.perm.trt, c(2, 3), data.frame) %>%
  rename(sig.id = X1, pathway = X2, null.delta = piece)

```

P-values

```{r}
signif(mat.p.trt, 2) %>% as.data.frame
```
Melt p-values table

```{r}
df.perm.p.trt <- as.data.frame(mat.p.trt) %>%
  rownames_to_column("sig.id") %>%
  tidyr::pivot_longer(-sig.id, values_to = "p.value", names_to = "pathway")
```

```{r}
df.perm.plot.trt <- left_join(df.orig.trt, df.perm.p.trt) %>%
  mutate(signif = p.value < .05, signif.range = as.ordered(cut(p.value, c(0, .05, .2, 1))))
  # mutate(signif = p.value < .05)
  # mutate(significance = ifelse(p.value < .05, "p<0.05", "p>=.05"))
```


P-values, added on top of original recap

```{r, fig.height=6, fig.width=9}
plt.p <- do.call(
  pheatmap, 
  c(list(mat = mat.orig.signif.trt, 
         color = heatmaps.pal(), 
         breaks = sym.breaks(mat.orig.trt)
         ), args.ph4)
)

ggsave(paste0(dir_out, "/IST_NASH_trt_withpvalue.png"), plot = plt.p)
```

Plot recap with significance (GSEA-like style)

```{r, fig.width=3.5, fig.height=2.25}
ggplot(df.perm.plot.trt,
       aes(y=sig.id, x=pathway, fill = total.delta.percent, 
           label = format(total.delta.percent, digits=0), color=signif.range)) +
  geom_tile(width=0.9, height=0.9, size=0.15) +
  ggplot2::geom_text(size = 1.1, color='black') +
  scale_colour_manual(values = c("black", "gray80", "white"), name = "Significance") +
  scale_fill_gradient2(low = "deepskyblue3", mid = "white", high = "darkorange1", name = "Recapitulation [%]") +
  # scale_y_reverse() +
  gg_def +
  # fix aspect ratio
  coord_fixed() +
  # fix legend
  guides(color = guide_legend(order = 2, override.aes = list(fill = "white")), 
         fill = guide_colorbar(order = 1)) +
  theme(
       # play with adding margins so that legend fits in plot
      legend.margin = margin(t = 0, r = 0, b = -.25, l = 0, unit = "cm"),
      plot.margin = margin(t = 1, r = .2, b = .2, l = .2, unit = "cm"),
      panel.grid = element_blank(),
    axis.line=element_blank(),
    panel.border=element_blank(),
    axis.ticks=element_blank(),
    text = element_text(size = 5, family = "Helvetica"),
    legend.key.size = unit(0.7,"line"),
    axis.text.x=element_text(angle=90, hjust=1, vjust=0.5),
    axis.title=element_blank())

ggsave(file = paste0(dir_out, '/main_nash_trt_heatmap_extended.pdf'), width=95, height=75, units='mm')
```

Plot reduced selection

```{r, fig.width=3.5, fig.height=2.25}
subset(df.perm.plot.trt, pathway %in% conf$analysis_nash_pathwayselection) %>%
  mutate(sig.id = factor(sig.id, levels = sig.trt), 
         pathway = factor(pathway, levels = conf$analysis_nash_pathwayselection)) %>%
  ggplot(aes(y=sig.id, x=pathway, fill = total.delta.percent, 
           label = format(total.delta.percent, digits=0), color=signif.range)) +
  geom_tile(width=0.9, height=0.9, size=0.15) +
  ggplot2::geom_text(size = 1.1, color='black') +
  scale_colour_manual(values = c("black", "gray80", "white"), name = "Significance") +
  scale_fill_gradient2(low = "deepskyblue3", mid = "white", high = "darkorange1", name = "Recapitulation [%]") +
  # scale_y_reverse() +
  gg_def +
      # fix aspect ratio
  coord_fixed() +
  # fix legend
  guides(color = guide_legend(order = 2, override.aes = list(fill = "white")), 
         fill = guide_colorbar(order = 1)) +
  theme(
       # play with adding margins so that legend fits in plot
      legend.margin = margin(t = 0, r = 0, b = -.25, l = 0, unit = "cm"),
      plot.margin = margin(t = 1, r = .2, b = .2, l = .2, unit = "cm"),
      panel.grid = element_blank(),
    axis.line=element_blank(),
    panel.border=element_blank(),
    axis.ticks=element_blank(),
    text = element_text(size = 5, family = "Helvetica"),
    legend.key.size = unit(0.7,"line"),
    axis.text.x=element_text(angle=90, hjust=1, vjust=0.5),
    axis.title=element_blank())

ggsave(file = paste0(dir_out, '/main_nash_trt_heatmap.pdf'), width=80, height=80, units='mm')
```

Plot distribution histograms

```{r, fig.width=12, fig.height=6}
list.path.plot <- list.path

ggplot() +
  geom_vline(xintercept = 0, colour = "gray50") +
  geom_vline(aes(xintercept = total.delta.percent, colour = signif.range), 
             lwd = .5, lty = 5, 
             data = subset(df.perm.plot.trt, pathway %in% list.path.plot)) +
  geom_histogram(aes(x = null.delta), 
                 data = subset(df.perm.distrib.trt, pathway %in% list.path.plot)) +
  facet_grid(sig.id~pathway, scales = "free") +
  scale_color_manual(values = c("indianred1", "darkgoldenrod2", "gray70")) +
  xlab("Recapitulation (null distribution, and observed statistic in dashed line)") +
  ylab("Counts (null distribution)") +
  theme_bw() +
  theme(strip.text.y = element_text(angle = 0), 
        strip.text.x = element_text(angle = 90))

ggsave(file = paste0(dir_out, '/main_nash_trt_histograms.pdf'))
```


## Export data objects

```{r}
saveRDS(ist.res.trt, paste0(dir_out, "/ist_res_trt.rds"))
save(mat.orig.trt, mat.p.trt, array.perm.trt, df.orig.trt, df.perm.distrib.trt, 
     df.perm.plot.trt, file = paste0(dir_out, "/null_models_trt.RData"))
```


# Reproducibility

```{r}
date()
```

```{r}
sessionInfo()
```